---
title: "Random Forest"
author: "Syed Jafar Raza Rizvi <br> NSID: cfr954 <br> Student ID: 11344782"
output:
  html_document:
    keep_md: true
date: "2025-02-11"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(visdat)
library(naniar)
library(mice)
library(bookdown)
library(VIM)
library(sjPlot)
library(C50)
library(finalfit)
library(knitr)
library(gtsummary)
library(mlbench)
library(vip)
library(rsample)
library(rpart.plot)
library(tune)
library(recipes)
library(yardstick)
library(parsnip)
library(glmnet)
library(themis)
library(microbenchmark)
library(arsenal)
library(workflows) 
library(dials)
```

1. **Dataset Exploration**  
   - Load and examine the **Can Path Student Dataset**. 
```{r}
data<-read.csv("mice_all_imp.csv")
data <- data %>% mutate_at(3, factor)
data <- data %>% mutate_at(5:6, factor)
data <- data %>% mutate_at(8:12, factor)
data <- data %>% mutate_at(15:81, factor)
data <- data %>% mutate_at(83:93, factor)

data$ID <- NULL
data$ADM_STUDY_ID <- NULL
```
   - Conduct an initial exploratory data analysis (EDA) to understand key features and relationships.  
We are going to use Diabatic as a predictor in the model. 

Diabctic (DIS_DIAB_EVER)  has 3 catagories:
0 - "Never had diabetes"
1-  "Ever had diabetes"
2 - "Presumed-Never had diabetes"

let see the frequency distribution of diabetic disease

```{r}
table(data$DIS_DIAB_EVER)
```
Let’s recode the variable DIS_DIAB_EVER into Ever had diabetes as Yes (1) , and Never had diabetes and presumed-Never had diabetes as No (0).
```{r}
data <- data %>%
	mutate(diabetes = case_when(
		DIS_DIAB_EVER == 0 ~ 0,
		DIS_DIAB_EVER == 1 ~ 1,
		DIS_DIAB_EVER == 2 ~ 0)) %>%
		mutate(diabetes = as.factor(diabetes))
table(data$DIS_DIAB_EVER, data$diabetes)
data$DIS_DIAB_EVER <- NULL
```
```{r}
table <- tableby(~diabetes   
                            , data=data)
knitr::kable(summary(table), caption = "Table 1 : Frequency distribution", booktabs = TRUE, "pipe")

ggplot(data) + 
    geom_bar(aes(diabetes))
```


From the table and graph we can say that, 7.6% of the study population has diabetes (1), while the majority (92.4%) does not (0). So we can say that the prevalence of diabetes is low in the dataset. we will use Synthetic Minority Over-sampling Technique (SMOTE) to incease diabetes cases.

   - Handle missing data, outliers, or inconsistencies if necessary.  
   
The dataset has been cleaned by using Multivariate Imputation by Chained Equations method.
<!-- We have selected following variable for our analysis  -->

<!-- SDC_AGE_CALC: Participant's age at questionnaire completion, calculated from birth date and date of questionnaire completion. -->
<!-- SDC_EDU_LEVEL: Highest level of education completed by the participant. -->
<!-- PM_BMI_SR: Body Mass Index. -->
<!-- WRK_FULL_TIME: Full-time employed. -->
<!-- SMK_CIG_EVER: Ever smoked 100 cigarettes or more. -->
<!-- SDC_INCOME: Household income before taxes -->
<!-- PA_TOTAL_SHORT: Total physical activity MET-minutes/week -->
<!-- HS_ROUTINE_VISIT_EVER: Ever had routine medical check-up. -->
<!-- PSE_ADULT_WRK_DURATION: Duration of passive smoking exposure during adulthood at work. -->
<!-- DIS_RESP_SLEEP_APNEA_EVER: Occurrence of sleep apnea. -->
<!-- SDC_EDU_LEVEL_AGE: Age highest level education completed. -->
<!-- SDC_GENDER: Gender of the participant. -->

<!-- ```{r} -->
<!-- data_small <- data %>%  -->
<!-- select (diabetes, SDC_AGE_CALC, SDC_EDU_LEVEL, PM_BMI_SR, WRK_FULL_TIME, SMK_CIG_EVER, SDC_INCOME, PA_TOTAL_SHORT, HS_ROUTINE_VISIT_EVER, PSE_ADULT_WRK_DURATION, DIS_RESP_SLEEP_APNEA_EVER, SDC_EDU_LEVEL_AGE, SDC_GENDER) -->
<!-- ``` -->


2. **Baseline Random Forest Model**  
   - Split the dataset into training and testing sets.  
```{r}
set.seed(786)


cv_split <- initial_split(data, 
                            strata = diabetes, 
                            prop = 0.70)
```
We have split the main dataset into two parts: the training dataset, where we put 70 percent of all observations, and the test dataset, where we put 30 percent. 
```{r}
train_data <- training(cv_split)


test_data  <- testing(cv_split)

```
### Set the number of cores on your computer
```{r}
cores <- parallel::detectCores()
cores
```
- Implement a baseline Random Forest model with default parameters.  
```{r}
rf_model_base <- rand_forest() %>% 
              set_engine("ranger", num.threads = cores) %>% 
              set_mode("classification")
```
## recipe bulding

```{r}
diabetes_recipe_base <- 
  recipe(diabetes ~ ., data = train_data) %>% 
  step_smotenc(diabetes, over_ratio = 0.9) %>%
  step_zv(all_predictors()) 
```

## workflow building
```{r}
library(recipes)
diabetes_workflow_base <- 
        workflow() %>% 
        add_model(rf_model_base) %>% 
        add_recipe(diabetes_recipe_base)

diabetes_workflow_base
```
## Fit the model 

```{r}
set.seed(100)

diabetes_workflow_fit_base <- diabetes_workflow_base %>% 
  fit(train_data)

diabetes_workflow_fit_base
summary(diabetes_workflow_fit_base)
```

## Test the trained model


```{r}
pred_diabetes_base_class <- predict(diabetes_workflow_fit_base,
                      new_data = test_data,
                      type = "class")
table(pred_diabetes_base_class$.pred_class)
```
```{r}
pred_diabetes_base_prob <- predict(diabetes_workflow_fit_base,
                      new_data = test_data,
                      type = "prob")
head(pred_diabetes_base_prob)
```
```{r}
diabetes_results_base <- test_data %>%
  select(diabetes) %>%
  bind_cols(pred_diabetes_base_class, pred_diabetes_base_prob)

head(diabetes_results_base)
```
   - Evaluate the model using appropriate metrics (e.g., accuracy, precision, recall, F1-score, or ROC-AUC).  
   
## Confusion Matrix
```{r}
conf_mat(diabetes_results_base, truth = diabetes,
         estimate = .pred_class)
```
## Accuracy

```{r}
accuracy_Based<-accuracy(diabetes_results_base, truth = diabetes,
         estimate = .pred_class)
accuracy_Based
```
## Sensitivity

```{r}
sensitivity_based<-sens(diabetes_results_base, truth = diabetes,
         estimate = .pred_class)
sensitivity_based
```

## Specificity

```{r}
specificity_based<-spec(diabetes_results_base, truth = diabetes,
         estimate = .pred_class)
specificity_based
```

## F1 Score

```{r}
f1_score_based<-f_meas(diabetes_results_base, truth = diabetes,
         estimate = .pred_class)
f1_score_based
```
#### Precision

```{r}
precision_based  <- precision(diabetes_results_base, truth = diabetes,
         estimate = .pred_class) 

precision_based
```

## ROC Curve

```{r}
roc_auc(diabetes_results_base,
        truth = diabetes,
        .pred_0)

roc_curve <- diabetes_results_base %>%
  roc_curve(truth = diabetes, .pred_0) %>%
  autoplot()

plot(roc_curve)
```
The ROC-AUC score of 0.7556 indicates that the model has moderate to good discriminative ability in distinguishing between positive and negative cases. It has a 75.56% chance of correctly ranking a positive instance higher than a negative one.

3. **Hyperparameter Tuning**  

# Build the model object 
```{r}
rf_model_tune <- rand_forest(mtry = tune(), min_n = tune(), trees = 100) %>% 
              set_engine("ranger", num.threads = cores) %>% 
              set_mode("classification")
```

# Recipe

```{r}
diabetes_recipe_tune <- 
  recipe(diabetes ~ ., data = train_data) %>% 
   step_smotenc(diabetes, over_ratio = 0.9) %>%
  step_zv(all_predictors()) ### Remove columns from the data when the training set data have a single value. Zero variance predictor
```

# Create a workflow

```{r}
diabetes_workflow_tune <- 
        workflow() %>% 
        add_model(rf_model_tune) %>% 
        add_recipe(diabetes_recipe_tune)

diabetes_workflow_tune
```

## fit model

```{r}
set.seed(100)

folds <- vfold_cv(train_data, v = 10) 

rf_grid <- grid_regular(
              mtry(range = c(1, 30)),
              min_n(range = c(5, 50)),
              levels = 5  
            )

rf_grid
```

```{r}
diabetes_fit_tune <- tune_grid(
                diabetes_workflow_tune,
                resamples = folds,
                grid = rf_grid, 
                control = control_resamples(save_pred = TRUE, 
                                                  verbose = FALSE))

diabetes_fit_tune
```
Evaluatinng the result


```{r}
diabetes_fit_tune %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
    values_to = "value",
    names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")

diabetes_fit_tune %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point() +
  labs(y = "Accuracy")
```
from the plot, the best tuned model have mtry around 8 and min_n of around 5 and trees of 100.
# Collect the results
```{r}
rf_best_tune <- 
  diabetes_fit_tune %>% 
  select_best(metric = "accuracy")

rf_best_tune
```
The best model is Preprocessor1_Model02 with mtry 8, min_n 5 and trees 100.

### Finalizing the best model 

```{r}
final_rf_model <- finalize_model(rf_model_tune, rf_best_tune)
```

### Creating final workflow
 
```{r}
rf_workflow_final <- workflow() %>% 
  add_model(final_rf_model) %>% 
  add_recipe(diabetes_recipe_tune) 
```

### Training the final Model 

```{r}
rf_workflow_final_fit <- rf_workflow_final %>% 
  fit(train_data)

rf_workflow_final_fit
```

### Prediciting test data 

#### Class Prediction

```{r}
final_model_predict <- predict(rf_workflow_final_fit,new_data = test_data) %>% 
  bind_cols(test_data)

final_model_predict<- final_model_predict  %>%
  select(.pred_class, diabetes)

final_model_predict %>% 
  head()
```



## Test the trained model


```{r}
pred_diabetes_base_class <- predict(final_model_predict,
                      new_data = test_data,
                      type = "class")
table(pred_diabetes_base_class$.pred_class)
```
```{r}
pred_diabetes_base_prob <- predict(final_model_predict,
                      new_data = test_data,
                      type = "prob")
head(pred_diabetes_base_prob)
```
```{r}
diabetes_results_base <- test_data %>%
  select(diabetes) %>%
  bind_cols(pred_diabetes_base_class, pred_diabetes_base_prob)

head(diabetes_results_base)
```
   - Evaluate the model using appropriate metrics (e.g., accuracy, precision, recall, F1-score, or ROC-AUC).  
   
## Confusion Matrix
```{r}
conf_mat(final_model_predict, truth = diabetes,
         estimate = .pred_class)
```
## Accuracy

```{r}
accuracy_tuned<-accuracy(final_model_predict, truth = diabetes,
         estimate = .pred_class)
accuracy_tuned
```
## Sensitivity

```{r}
sensitivity_tuned<-sens(final_model_predict, truth = diabetes,
         estimate = .pred_class)
sensitivity_tuned
```

## Specificity

```{r}
specificity_tuned<-spec(final_model_predict, truth = diabetes,
         estimate = .pred_class)
specificity_tuned
```

## F1 Score

```{r}
f1_score_tuned<-f_meas(final_model_predict, truth = diabetes,
         estimate = .pred_class)
f1_score_tuned
```
#### Precision

```{r}
precision_tuned  <- precision(final_model_predict, truth = diabetes,
         estimate = .pred_class) 

precision_tuned
```
#### Combine metrics into a data frame

```{r}
tuned_metrics <- data.frame(
  accuracy_tuned = accuracy_tuned$.estimate,
  sensitivity_tuned = sensitivity_tuned$.estimate,
  specificity_tuned = specificity_tuned$.estimate,
  precision_tuned= precision_tuned$.estimate,
  f_score_tuned = f1_score_tuned$.estimate
)

tuned_metrics %>% 
  head()
```

#### Changing shape of dataframe 

```{r}
tuned_metrics_long <- tuned_metrics %>% 
  pivot_longer(cols = accuracy_tuned:f_score_tuned,
               names_to = "metrics",
               values_to = "estimates")

tuned_metrics_long
```




## ROC Curve

```{r}
roc_auc(final_model_predict,
        truth = diabetes,
        .pred_class)

roc_curve <- final_model_predict %>%
  roc_curve(truth = diabetes, .pred_class) %>%
  autoplot()

plot(roc_curve)
```




4. **Model Comparisons**  
   - Compare the tuned Random Forest model against:  
     - The baseline Random Forest model.  
   - Analyze differences in performance using quantitative metrics and discuss possible reasons for the results.  

5. **Feature Importance Analysis**  
   - Extract and interpret feature importance scores from the Random Forest model.  
   - Visualize the top contributing features and discuss their relevance to the dataset.  
   
   
   