---
title: "Independent Analysis"
author: "Syed Jafar Raza Rizvi <br> NSID: cfr954 <br> Student ID: 11344782"
output:
  html_document:
    keep_md: true
date: "2025-02-11"
---

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)
library(sjPlot)
library(finalfit)
library(knitr)
library(gtsummary)
library(mlbench)
library(kernlab)
library(vip)
library(rsample)
library(tune)
library(recipes)
library(yardstick)
library(parsnip)
library(glmnet)
library(themis)
library(microbenchmark)
library(naniar)
```

We are using the imputed (ie. no missing data) version of the synthetic cohort of patients who have no prior history of AF and had a baseline ECG performed between January 2010 and January 2023 , followed by a minimum follow-up of 12 months. 

### Research Questions:
Development and Validation of a New-Onset Atrial Fibrillation Prediction Model?

In this study, we utilized a comprehensive dataset comprising 12-lead electrocardiogram (ECG) measurements, laboratory results, medical history, and demographic variables. Missing data were observed in the laboratory variables, which were addressed using Multiple Imputation by Chained Equations (MICE), a robust statistical method for handling missing values. The imputation process was conducted with 20 imputations and 5 iterations to ensure convergence and stability of the estimates. This approach preserves statistical power and reduces bias while accounting for the uncertainty inherent in missing data.

## Reading data 
```{r}
data_synt1 <- read_csv("data_synt1_mice.csv")
```

```{r}
get_unique_values <- function(df, vars_to_exclude = NULL) {
  # Initialize results dataframe
  result <- data.frame(
    Variable_Name = character(),
    Unique_Values = character(),
    Is_Boolean = logical(),
    # Value_Count = integer(),
    stringsAsFactors = FALSE
  )
  
  # Get columns to analyze (excluding specified vars)
  cols_to_analyze <- setdiff(names(df), vars_to_exclude)
  
  for (col in cols_to_analyze) {
    # Get complete unique values (including NA)
    unique_vals <- unique(df[[col]])
    na_present <- any(is.na(unique_vals))
    na_sum<-sum(is.na(df[[col]]))
    na_sum_per<-((sum(is.na(df[[col]])))/length(df[[col]]))*100
    
    # Remove NA for boolean detection
    clean_vals <- unique_vals[!is.na(unique_vals)]
    val_count <- length(clean_vals)
    
    # Detect boolean variables (0/1 or 1/2 numeric)
    is_boolean <- if (val_count <= 2 && is.numeric(df[[col]])) {
      all(clean_vals %in% c(0, 1)) || all(clean_vals %in% c(1, 2))
    } else {
      FALSE
    }
    
    # Prepare values for display
    if (na_present) {
      display_vals <- c(sort(clean_vals), NA)
    } else {
      display_vals <- sort(clean_vals)
    }
    
    # Convert to character representation
    char_vals <- sapply(display_vals, function(x) {
      if (is.na(x)) "NA" else as.character(x)
    })
    
    result <- rbind(result, data.frame(
      Variable_Name = col,
      Unique_Values = paste(char_vals, collapse = ", "),
      Is_Boolean = is_boolean,
      Value_Count = val_count,
      missing=na_sum,
      missing_per=na_sum_per,
      stringsAsFactors = FALSE
    ))
  }
  
  return(result)
}
unique_values <- get_unique_values(data_synt1)
table(unique_values$Is_Boolean)
 
unique_values<- unique_values %>%
  filter(Variable_Name!="patient_id")
unique_values1<- unique_values %>%
  filter(Value_Count==1)
print(unique_values1$Variable_Name)
```
those variable have only outcome. so we are going to drop those variables.

```{r}
data_synt1 <- data_synt1 %>% 
  select(-any_of(unique_values1$Variable_Name))
```

for this analysis we only consider the ECG variable which are 

Heart rate from index 12 lead ECG [ecg_resting_hr]
PR interval duration from index 12 lead ECG [ecg_resting_pr]
QRS complex duration from index 12 lead ECG [ecg_resting_qrs]
Corrected QT interval  from index 12 lead ECG [ecg_resting_qtc]
Atrial fibrillation in index 12 lead ECG [ecg_resting_afib]
Paced rhythm in index 12 lead ECG [ecg_resting_paced]
Bigeminy in index 12 lead ECG [ecg_resting_bigeminy]
Left bundle branch block in index 12 lead ECG [ecg_resting_LBBB]
Right bundle branch block in index 12 lead ECG [ecg_resting_RBBB]
Incomplete left bundle branch block in index 12 lead ECG [ecg_resting_incomplete_LBBB]
Incomplete right bundle branch block in index 12 lead ECG [ecg_resting_incomplete_RBBB]
Left anterior fascicular block in index 12 lead ECG [ecg_resting_LAFB]
Left posterior fascicular block in index 12 lead ECG [ecg_resting_LPFB]
Bifascicular block in index 12 lead ECG [ecg_resting_bifascicular_block]
Trifascicular block in index 12 lead ECG [ecg_resting_trifascicular_block]
Intra-ventricular conduction delay in  index 12 lead ECG [ecg_resting_intraventricular_conduction_delay]
Chronological age at time of referenced index ECG, rounded down to nearest whole number [demographics_age_index_ecg]
Sex assigned at birth [demographics_birth_sex]
ICD-10 coding of hypertension in either DAD or NACRS databases at any time prior to or 6 months after date of index 12 lead ECG [hypertension_icd10]
Documented presence of hyperglycaemic state inclusive of Type 1 diabetes, Type 2 diabetes, Pre-diabetes, or Impaired fasting glucose (treated or untreated) by any of: labs, medications, or ICD-10,  at any time prior to or 6 months after date of index 12 lead ECG [diabetes_combined]
Documented presence of dyslipidemia (treated or untreated) by any of: labs, medications, or ICD-10,  at any time prior to or 6 months after date of index 12 lead ECG [dyslipidemia_combined]
ICD-10 coded presence of dilated cardiomyopathy at any time prior to or 6 months after date of index 12 lead ECG [dcm_icd10]
ICD-10 coded presence of hypertrophic cardiomyopathy at any time prior to or 6 months after date of index 12 lead ECG [hcm_icd10]
ICD-10 coded presence of acute myocarditis at any time prior to date of index 12 lead ECG [myocarditis_icd10_prior]
ICD-10 coded presence of acute pericarditis at any time prior todate of index 12 lead ECG [pericarditis_icd10_prior]
ICD-10 coded presence of aortic dilatation/aneurysm at any time prior to or 6 months after date of index 12 lead ECG [aortic_aneurysm_icd10]
ICD-10 coded presence of aortic dissection at any time prior to date of index 12 lead ECG [aortic_dissection_icd10_prior]
ICD-10 coded presence of pulmonary hypertension at any time prior to or 6 months after date of index 12 lead ECG [pulmonary_htn_icd10]
ICD-10 coded presence of amyloidosis at any time prior to or 6 months after date of index 12 lead ECG [amyloid_icd10]
ICD-10 coded presence of chronic obstructive airway disease (COPD) at any time prior to or 6 months after date of index 12 lead ECG [copd_icd10]
ICD-10 coded presence of hyperthyroidism at any time prior to or 6 months after date of index 12 lead ECG [hyperthyroid_icd10]
ICD-10 coded presence of hypothyroidism at any time prior to or 6 months after date of index 12 lead ECG [hypothyroid_icd10]
ICD-10 coded presence of rheumatoid arthritis at any time prior to or 6 months after date of index 12 lead ECG [rheumatoid_arthritis_icd10]
ICD-10 coded presence of systemic lupus erythematosus (SLE) at any time prior to or 6 months after date of index 12 lead ECG [sle_icd10]
ICD-10 coded presence of sarcoidosis at any time prior to or 6 months after date of index 12 lead ECG [sarcoid_icd10]
ICD-10 coded presence of cancer at any time prior to or 6 months after date of index 12 lead ECG [cancer_any_icd10]

```{r}
data_synt1 <- data_synt1 %>%
  dplyr::select(ecg_resting_hr,
ecg_resting_pr,
ecg_resting_qrs,
ecg_resting_qtc,
ecg_resting_paced,
ecg_resting_bigeminy,
ecg_resting_LBBB,
ecg_resting_RBBB,
ecg_resting_incomplete_LBBB,
ecg_resting_incomplete_RBBB,
ecg_resting_LAFB,
ecg_resting_LPFB,
ecg_resting_bifascicular_block,
ecg_resting_trifascicular_block,
ecg_resting_intraventricular_conduction_delay,
outcome_afib_aflutter_new_post,
demographics_age_index_ecg,
demographics_birth_sex,
hypertension_icd10,
diabetes_combined,
dyslipidemia_combined,
dcm_icd10,
hcm_icd10,
myocarditis_icd10_prior,
pericarditis_icd10_prior,
aortic_aneurysm_icd10,
aortic_dissection_icd10_prior,
pulmonary_htn_icd10,
amyloid_icd10,
copd_icd10,
hyperthyroid_icd10,
hypothyroid_icd10,
rheumatoid_arthritis_icd10,
sle_icd10,
sarcoid_icd10,
cancer_any_icd10)
```

**Missing Data Information**

```{r}
missing_table <- miss_var_summary(data_synt1)
missing_table
```

```{r}
factor_vars <- unique_values %>% 
  filter(Is_Boolean== TRUE) %>% 
  pull(Variable_Name)
data_synt1 <- data_synt1 %>% 
  mutate(across(any_of(factor_vars), as.factor))
data_synt1$patient_id<-NULL
```

```{r}
table(data_synt1$outcome_afib_aflutter_new_post)
```

### Prepare the data split and cross validation folds

```{r}
set.seed(10)

#### Cross Validation Split
cv_split <- initial_validation_split(data_synt1, 
                            strata = outcome_afib_aflutter_new_post, 
                            prop = c(0.70, 0.20))

# Create data frames for the two sets:
train_data <- training(cv_split)
table(train_data$outcome_afib_aflutter_new_post)
```
```{r}
test_data  <- testing(cv_split)
table(test_data$outcome_afib_aflutter_new_post)
```
### V folds

```{r}
folds <- vfold_cv(training(cv_split), v = 5, strata = outcome_afib_aflutter_new_post)
```



```{r}
all_recipe <- 
  recipe(outcome_afib_aflutter_new_post ~ ., data = train_data) %>%
  step_smotenc(outcome_afib_aflutter_new_post, over_ratio = 0.9) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors(), -all_outcomes()) %>%
  step_normalize(all_numeric_predictors())
```

### logistic regression 

```{r}
logistic_model <- logistic_reg(penalty = tune(), mixture = tune(),
                                mode = "classification",
                                engine = "glmnet"
                               )

logistic_workflow <- workflow() %>% 
          add_model(logistic_model) %>% 
          add_recipe(all_recipe) %>% 
          tune_grid(resamples = folds,
                    control = control_grid(save_pred = TRUE, 
                                            verbose = FALSE)) ## Edit for running live

collect_metrics(logistic_workflow) 
```


```{r}
show_best(logistic_workflow, metric='accuracy', n=5)  # only show the results for the best 5 models
```
```{r}
autoplot(logistic_workflow, metric = 'accuracy') 
```
### Random Forest 

```{r}
cores <- parallel::detectCores()
cores
```

```{r}
rf_model <- rand_forest(mtry = tune(), min_n = tune(), trees = tune()) %>% 
              set_engine("ranger", num.threads = cores) %>% 
              set_mode("classification")

rf_workflow <- workflow() %>% 
          add_model(rf_model) %>% 
          add_recipe(all_recipe) %>% 
          tune_grid(resamples = folds,
                    control = control_grid(save_pred = TRUE, 
                                            verbose = FALSE)) ## Edit for running live
```

```{r}
collect_metrics(rf_workflow) 
```

```{r}
show_best(rf_workflow, metric='accuracy', n=5)  # only show the results for the best 5 models
```

```{r}
autoplot(rf_workflow, metric = 'accuracy') 
```

#### Support Vector Machine

```{r}
svm_model <- svm_rbf(cost = tune(), rbf_sigma = tune()) %>% 
  set_mode("classification") %>%
  set_engine("kernlab")

svm_workflow <- workflow() %>% 
          add_model(svm_model) %>% 
          add_recipe(all_recipe) %>% 
          tune_grid(resamples = folds,
                    control = control_grid(save_pred = FALSE, 
                                            verbose = FALSE)) ## Edit for running live
```

```{r}
collect_metrics(svm_workflow) 
```

```{r}
show_best(svm_workflow, metric='accuracy', n=5)  # only show the results for the best 5 models
```

```{r}
autoplot(svm_workflow, metric = 'accuracy') 
```

### Best Model 

## logistic Regression

```{r}
logistic_best <- 
  logistic_workflow %>% 
  select_best(metric = "brier_class")

logistic_final_model <- finalize_model(
                          logistic_model,
                          logistic_best
                          )
logistic_final_model
```

```{r}
final_logistic_workflow <- workflow() %>%
                      add_recipe(all_recipe) %>%
                      add_model(logistic_final_model)

final_logistic_results <- final_logistic_workflow %>%
                    last_fit(cv_split)

lr_results <- final_logistic_results %>% collect_metrics()
```


## Random Forest

```{r}
rf_best <- 
  rf_workflow %>% 
  select_best(metric = "brier_class")

rf_final_model <- finalize_model(
                          rf_model,
                          rf_best
                          )
rf_final_model
```

```{r}
final_rf_workflow <- workflow() %>%
                      add_recipe(all_recipe) %>%
                      add_model(rf_final_model)

final_rf_results <- final_rf_workflow %>%
                    last_fit(cv_split)

rf_results <- final_rf_results %>% collect_metrics()
```

## Support Vector Machine

```{r}
svm_best <- 
  svm_workflow %>% 
  select_best(metric = "brier_class")

svm_final_model <- finalize_model(
                          svm_model,
                          svm_best
                          )
svm_final_model
```

```{r}
final_svm_workflow <- workflow() %>%
                      add_recipe(all_recipe) %>%
                      add_model(svm_final_model)

final_svm_results <- final_svm_workflow %>%
                      last_fit(cv_split)

svm_results <- final_svm_results %>% collect_metrics()
```

### Final Resul

## Logistic Regression 

```{r}
kable(lr_results)
```

## Random Forest 

```{r}
kable(rf_results)
```

## SVM 

```{r}
kable(svm_results)
```


